---
title: "PredictionAssignment"
author: "VRS"
---

#Synopsis

##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Data

The training and test data can be obtained from these locations:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The original source of the data is:
http://groupware.les.inf.puc-rio.br/har.

#Read the Data

Let us first load the datasets.

```{r, warning=FALSE}
#Load the packages
library(rpart)
library(randomForest)
library(caret)

#Download data
if(!file.exists("Data")) {
  dir.create("Data")
}
if(!file.exists("./Data/pml-training.csv")) {
  trainingUrl<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
  download.file(trainingUrl, destfile = "./Data/pml-training.csv")
}
if(!file.exists("./Data/pml-testing.csv")) {
  testingUrl<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
  download.file(testingUrl, destfile = "./Data/pml-testing.csv")
}

#Read data
training<-read.csv("./Data/pml-training.csv")
validate<-read.csv("./Data/pml-testing.csv")
```

#Data Preprocessing
#Clean the dataset
Let us remove entries that have "NA".

```{r}
training<-training[,(colSums(is.na(training))==0)]
```
Features unrelated to the problem at hand, such as X, user_name, raw_timestamp_part_1, etc. are removed.

```{r}
index <-  which(names(training) %in% c('X', 'user_name', 'raw_timestamp_part_1', 'raw_timestamp_part_2', 'cvtd_timestamp', 'new_window','num_window'))
training <- training[, -index]
```

Some features have zero variance, which in turn implies that they will not provide any infomartion for the classification. Let us remove such features.

```{r}
trainNzv <- nearZeroVar(training, saveMetrics=TRUE)
training <- training[,trainNzv$nzv==FALSE]
```
# Data Partitioning
Let us partition the dataset into training and testing.

```{r}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
train <- training[inTrain,]
test <- training[-inTrain,]
```

#Model Building

Let us apply three models: Decision trees, generalized booseted regression and finally random forests.

```{r, cache=TRUE, warning=FALSE, message=FALSE}
#Decision trees
modelDecisionTrees<-train(classe~.,data=train, method="rpart")
modelDecisionTrees

#Generalized boosted regression
control<-trainControl(number=5)
modelGBM<-train(classe~.,data=train, method="gbm", trControl=control)
modelGBM

#Random forests
control<-trainControl(number=5)
modelRandomForests<-train(classe~.,data=train, method="rf", trControl=control)
modelRandomForests
```

#Model Testing

Let us test the trained models with the testing data.

```{r}
#Decision tree
predictDecisionTrees<-predict(modelDecisionTrees, test)
confusionMatrix(predictDecisionTrees, test$classe)

#Test the generalized boosted regression model
predictGBM<-predict(modelGBM,test)
confusionMatrix(predictGBM, test$classe)

#Test the random forests model
predictRandomForest<-predict(modelRandomForests,test)
confusionMatrix(predictRandomForest, test$classe)
```
It can be seen from the above results that Random Forests performed the best with an accuracy of 0.99, while decision trees performed the worst.

Hence we will choose Random Forests as our model.

#Model Validation

Let us now validate of results with the validation set.
```{r}
validation<-predict(modelRandomForests, validate)
validation
```

We can see that the validation yields satisfactory results.